Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-04-21 appendix-EPQ.tex
something
modify

A path of an LTS is a finite transition sequence $q_0\xrightarrow{\beta_1}q_1\overset{\beta_2}{\longrightarrow}\ldots\overset{\beta_k}{\longrightarrow}q_k$ for $k\geq 0$, where $q_0$ is the initial state of the LTS. A trace of an LTS is a finite sequence $\beta_1 \cdot \beta_2 \cdot \ldots \cdot \beta_k$, where $k \geq 0$ if there exists a path $q_0\overset{\beta_1}{\longrightarrow}q_1\overset{\beta_2}{\longrightarrow}\ldots\overset{\beta_k}{\longrightarrow}q_k$ of the LTS. Let $\seqPQ$ be the set of traces of $\textit{PQ}$. The following lemma states that $\seqPQ$ is indeed the set of sequences obtained by renaming sequences accepted by $\textit{Check-PQ-Seq}$.

$\textit{Check-PQ-Seq}(e)=\mathsf{true}$ iff $e\in \seqPQ$, for every data-differentiated sequential execution $e$..

@@

Let $\textit{SeqPQ}_f$ be the set of data-differentiated sequences, such that $e \in \textit{SeqPQ}_f$, if $\textit{Check-PQ-Seq}(e)=\mathsf{true}$. We need to prove that $\textit{SeqPQ}_f = \seqPQ_{\neq}$. Given $\Gamma\in \{\mathsf{EmptyRemove}, \mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$, let us use $l_2 \xrightarrow{\Gamma} l_1$ to mean that when we use $\textit{Check-PQ-Seq}$ to check $l_2$, we choose the branch of $\mathsf{Has\text{-}\Gamma}$ and finally recursively call $\textit{Check-PQ-Seq}$ to check $l_1$.

We prove $\textit{SeqPQ}_f \subseteq \seqPQ_{\neq}$ by induction.

@@

$\textit{Check-PQ-Conc}(e)=\mathsf{true}$ iff $e \sqsubseteq \seqPQ$, for every data-differentiated execution $e$.

@@

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] If $\mathsf{Has\text{-}EmptyRemoves}(e)$, then $l = u \cdot o \cdot v$ for some $o=\textit{rm}(\textit{empty})$. It is not hard to see that $\mathsf{EmptyRemove\text{-}Seq}(l,o)$. Let $e'$ be the projection of $e$ into operations of $u \cdot v$. It is easy to see that $e' \sqsubseteq u \cdot v$ and then $e' \sqsubseteq \seqPQ$. Then we start the next round and choose $e'$ to be the ``$e$ in the next round''.

\item[-] \item[-] If $\mathsf{Has\text{-}UnmatchedMaxPriority}(e)$, then there exists a maximal priority $p$ in $e$, such that $p$ has unmatched $\textit{put}$. Let $l = u \cdot \textit{put}(x,p) \cdot v$, where in $v$ there is no $\textit{put}$ with priority $p$. It is not hard to see that $\mathsf{UnmatchedMaxPriority\text{-}Seq}(l,x)$. Let $e'$ be the projection of $e$ into operations of $u \cdot v$. It is easy to see that $e' \sqsubseteq u \cdot v$ and then $e' \sqsubseteq \seqPQ$. Then we start the next round and choose $e'$ to be the ``$e$ in the next round''.

where $\mathsf{Has\text{-}\Gamma}(e)$ holds and .

The proof of the $\textit{if}$ direction is quite similar to the proof of $\seqPQ_{\neq} \subseteq \textit{SeqPQ}_f$ direction of Lemma \ref{lemma:EPQ rules and semantics} and is omitted here.

To prove the $\textit{only if}$ direction, given a data-differentiated execution $e$ and assume that $\textit{Check-PQ-Conc}(e)=\mathsf{true}$.

$e \sqsubseteq \seqPQ$, according to the definition of $\textit{PQ}$, we can always find

@@

To prove the $\textit{only if}$ direction, assume that $e \sqsubseteq l \in \seqPQ$. Given $e' = e \vert_{D}$ and $l' = l \vert_{D}$, it is easy to see that $e' \sqsubseteq l'$, and by Lemma \ref{lemma:EPQ is closed under projection}, we can see that $l' \in \seqPQ$. Then by Lemma \ref{lemma:EPQ as multi in MRpri for sequence} we know that for each $R \in \textit{last}(l')$, we have $l' \in \textit{MS}(R)$.

To prove the $\textit{if}$ direction, given $e_1 = e$, we generate sequence $e_2$ from $e_1$ as follows: Since $e_1 \in \textit{proj}(e)$, we know that for each $R_1 \in \textit{last}(e_1)$, we have $e_1 \sqsubseteq \textit{MS}(R_1)$. We choose an arbitrary $R_1$ in $\textit{last}(e_1)$,

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] If $R_1 = \seqPQ_3$: $e_2$ is generated from $e_1$ by erasing call and return of one $\textit{rm}(\textit{empty})$ operation.

\item[-] Else, if $R_1 = \seqPQ_2^{=}, \seqPQ_2^{>}, \seqPQ_1^{=}, \seqPQ_1^{>}$, and the witness is $x$: $e_2$ is generated from $e_1$ by erasing call and return of method event of item $x$.
\end{itemize}

Similarly, for each $i > 1$, we obtain $e_{\textit{i+1}}$ from $e_i$, until we obtain $e_m = \epsilon$ for some $m$. It is obvious that $e_m \sqsubseteq \seqPQ$. For $e_{\textit{m-1}}$, since

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] $e_m \sqsubseteq \seqPQ$,

\item[-] If $\textit{last}(e_{\textit{m-1}}) = R_{\textit{m-1}} \in \{ \seqPQ_1^{>}, \seqPQ_1^{=}, \seqPQ_2^{>}, \seqPQ_2^{=} \}$ and $e_{\textit{m-1}}$ matches $R_{\textit{m-1}}$ with witness $x$: We already know that $e_{\textit{m-1}} \sqsubseteq \textit{MS}(R_{\textit{m-1}})$, $e_m = e_{\textit{m-1}} \setminus x \sqsubseteq \seqPQ$, and by step-by-step linearizability of $\seqPQ$ (Lemma \ref{lemma:EPQ is step-by-step linearizability}), we can see that $e_{\textit{m-1}} \sqsubseteq \seqPQ$.

\item[-] If $\textit{last}(e_{\textit{m-1}}) = R_{\textit{m-1}} = \seqPQ_3$ and $o$ is a $\textit{rm}(\textit{empty})$ in $e_{\textit{m-1}}$ that is removed in the process of constructing $e_m$: We already know that $e_{\textit{m-1}} \sqsubseteq \textit{MS}(R_{\textit{m-1}})$, $e_m = e_{\textit{m-1}} \setminus o \sqsubseteq \seqPQ$, and by step-by-step linearizability of $\seqPQ$ (Lemma \ref{lemma:EPQ is step-by-step linearizability}), we can see that $e_{\textit{m-1}} \sqsubseteq \seqPQ$.
\end{itemize}

Therefore, we know that $e_{\textit{m-1}} \in \seqPQ$. Similarly, we can prove that $e_{\textit{m-2}},\ldots,e_1 = e \in \seqPQ$. \qed
\end {proof}

\subsection{Proof of Lemma \ref{lemma:EPQ as multi in MRpri for sequence}}

%\EPQasMultiInMRpriforSequence*

\begin{restatable}{lemma}{EPQasMultiInMRpriforSequence}
\label{lemma:EPQ as multi in MRpri for sequence}
Given a data-differentiated sequential execution $e$, $e \in \seqPQ$, if and only if, $\forall e' \in \textit{proj}(e)$ and $\forall R \in \textit{last}(e')$, we have $e' \in \textit{MS}(R)$.
\end{restatable}

The \textit{only if} direction is obvious and can be similarly proved as the $\seqPQ_s \subseteq \seqPQ$ direction of Lemma \ref{lemma:EPQ rules and semantics}.

To prove the $\textit{if}$ direction, we proceed as follows: From $e1=e$, we generate a sequence $e_2$ as follows:

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] If $\textit{PQ}_3 \in \textit{last}(e_1)$: Then we can see that $e_1$ contains at least one $\textit{rm}(\textit{empty})$. $e_2$ is generated from $e_1$ by erasing one $\textit{rm}(\textit{empty})$.

\item[-] Else, if $\textit{PQ}_2^{=} \in \textit{last}(e_1)$: Then we can see that one of the maximal priority of $e_1$ is unmatched $\textit{put}$ and (possibly) matched $\textit{put}$. Assume the set of the items of these unmatched $\textit{put}$ is $S$. $e_2$ is generated from $e_1$ by erasing one unmatched $\textit{put}$ which use the item last putted in $S$.

\item[-] Else, if $\textit{PQ}_2^{>} \in \textit{last}(e_1)$: Then we can see that one of the maximal priority in $e_1$ is unmatched $\textit{put}$. Assume the set of the items of these unmatched $\textit{put}$ is $S$. $e_2$ is generated from $e_1$ by erasing one unmatched $\textit{put}$ which use the item last putted in $S$.

\item[-] Else, if $\textit{PQ}_1^{=} \in \textit{last}(e_1)$: Then we can see that one of the maximal priority in $e_1$ is of more than one pair of matched $\textit{put}$. Assume the set of the items of these matched $\textit{put}$ is $S$. $e_2$ is generated from $e_1$ by erasing matched $\textit{put}$ and $\textit{rm}$ of the item which is last putted in $S$.

\item[-] Else, if $\textit{PQ}_1^{>} \in \textit{last}(e_1)$: Then we can see that one of the maximal priority in $e_1$ is of one pair of matched $\textit{put}$. $e_2$ is generated from $e_1$ by erasing this pair of matched $\textit{put}$ and $\textit{rm}$.
\end{itemize}

Similarly, for each $i > 1$, we obtain $e_{\textit{i+1}}$ from $e_i$, until we obtain $e_m = \epsilon$ for some $m$. It is obvious that $e_m \in \seqPQ$. For $e_{\textit{m-1}}$, since

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] $e_m \in \seqPQ$,

\item[-] By assumption, we know that $e_{\textit{m-1}} \in \textit{MS}(R_{\textit{m-1}})$, where $R_{\textit{m-1}} \in \textit{last}(e_{\textit{m-1}})$. This implies that the guard of $R_{\textit{m-1}}$ is satisfied.
\end{itemize}

Therefore, we know that $e_{\textit{m-1}} \in \seqPQ$. Similarly, we can prove that $e_{\textit{m-2}},\ldots,e_1 = e \in \seqPQ$. \qed
\end {proof}

@@

\EPQOneLarisCoRegular*

@@

\EPQOneEqualIsCoRegular*

>>>>

A path of an LTS is a finite transition sequence $q_0\xrightarrow{\beta_1}q_1\overset{\beta_2}{\longrightarrow}\ldots\overset{\beta_k}{\longrightarrow}q_k$ for $k\geq 0$, where $q_0$ is the initial state of the LTS. A trace of an LTS is a finite sequence $\beta_1 \cdot \beta_2 \cdot \ldots \cdot \beta_k$, where $k \geq 0$ if there exists a path $q_0\overset{\beta_1}{\longrightarrow}q_1\overset{\beta_2}{\longrightarrow}\ldots\overset{\beta_k}{\longrightarrow}q_k$ of the LTS. Let $\seqPQ$ be the set of traces of $\textit{PQ}$. The following lemma states that $\seqPQ$ is indeed the set of sequences obtained by renaming sequences accepted by $\textit{Check-PQ-Seq}$. Given $\Gamma\in \{\mathsf{EmptyRemove}, \mathsf{Unmatched-}$ $\mathsf{MaxPriority}, \mathsf{MatchedMaxPriority}\}$, let us use $l_2 \xrightarrow{\Gamma} l_1$ to mean that when we use $\textit{Check-PQ-Seq}$ to check $l_2$, we choose the branch of $\mathsf{Has\text{-}\Gamma}$ and finally recursively call $\textit{Check-PQ-Seq}$ to check $l_1$.

@@

Let $\textit{SeqPQ}_f$ be the set of data-differentiated sequences, such that $e \in \textit{SeqPQ}_f$, if $\textit{Check-PQ-}$ $\textit{Seq}(e)=\mathsf{true}$. We need to prove that $\textit{SeqPQ}_f = \seqPQ_{\neq}$. 
We prove $\textit{SeqPQ}_f \subseteq \seqPQ_{\neq}$ by induction.

@@

@@

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] If $\mathsf{Has\text{-}EmptyRemoves}(e)$ holds, then $l = u \cdot o \cdot v$ for some $o=\textit{rm}(\textit{empty})$. It is not hard to see that $\mathsf{EmptyRemove\text{-}Seq}(l,o)$ holds. Let $e'$ be the projection of $e$ into operations of $u \cdot v$. It is easy to see that $e' \sqsubseteq u \cdot v$ and then $e' \sqsubseteq \seqPQ$. Then we start the next round and choose $e'$ to be the ``$e$ in the next round''.

\item[-] If $\mathsf{Has\text{-}UnmatchedMaxPriority}(e)$ holds, then there exists a maximal priority $p$ in $e$, such that $p$ has unmatched $\textit{put}$. Let $l = u \cdot \textit{put}(x,p) \cdot v$, where in $v$ there is no $\textit{put}$ with priority $p$. It is not hard to see that $\mathsf{UnmatchedMaxPriority\text{-}Seq}(l,x)$ holds. Let $e'$ be the projection of $e$ into operations of $u \cdot v$. It is easy to see that $e' \sqsubseteq u \cdot v$ and then $e' \sqsubseteq \seqPQ$. Then we start the next round and choose $e'$ to be the ``$e$ in the next round''.

\item[-] If $\mathsf{Has\text{-}MatchedMaxPriority}(e)$ holds, then there exists a maximal priority $p$ in $e$, such that $p$ has only matched $\textit{put}$. Let $l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w$, where in $v \cdot w$ there is no $\textit{put}$ with priority $p$. It is not hard to see that $\mathsf{MatchedMaxPriority\text{-}Seq}(l,x)$ holds. Let $e'$ be the projection of $e$ into operations of $u \cdot v \cdot w$. It is easy to see that $e' \sqsubseteq u \cdot v \cdot w$ and then $e' \sqsubseteq \seqPQ$. Then we start the next round and choose $e'$ to be the ``$e$ in the next round''. 
\end{itemize}

The loop terminates when at some round $e=\epsilon$. This process ensures that $\textit{Check-PQ-Conc}(e)=\mathsf{true}$.

To prove the $\textit{only if}$ direction, given a data-differentiated execution $e$ and assume that $\textit{Check-PQ-}$ $\textit{Conc}(e)=\mathsf{true}$. Then we have $e \xrightarrow{\Gamma_1} e_1 \ldots \xrightarrow{\Gamma_m} e_m$ with $e_0=e$ and $e_m=\epsilon$. Assume that for each $i$, we found $\mathsf{\Gamma\text{-}Seq}(e_i,\alpha_i)$ holds for some $\alpha_i$. From $\mathsf{\Gamma\text{-}Seq}(e_{\textit{m-1}},\alpha_{\textit{m-1}})$ holds and $e_m = e_{\textit{m-1}} \setminus \alpha_{\textit{m-1}} \sqsubseteq \seqPQ$, by Lemma \ref{lemma:EPQ is step-by-step linearizability}, we can see that $e_{\textit{m-1}} \sqsubseteq \seqPQ$. Similarly, we can prove that $e_{\textit{m-2}},\ldots,e_0=e \sqsubseteq \seqPQ$. \qed
\end {proof}

@@

To prove the $\textit{only if}$ direction, assume that $e \sqsubseteq l \in \seqPQ$. Given $e' = e \vert_{D}$ and $l' = l \vert_{D}$, it is easy to see that $e' \sqsubseteq l'$, and by Lemma \ref{lem:closure_proj}, we can see that $l' \in \seqPQ$. Then it is obvious that $\textit{Check-PQ-Conc-NonRec}(e')$ returns $\mathsf{true}$.

To prove the $\textit{if}$ direction, assume that for each $e' \in \textit{proj}(e)$, $\textit{Check-PQ-Conc-NonRec}(e')$ returns $\mathsf{true}$. Then similarly as by Lemma \ref{lemma:con-check-EPQ is correct}, we can implies that $\textit{Check-PQ-Conc}(e)=\mathsf{true}$, and then by Lemma \ref{lemma:con-check-EPQ is correct}, we know that $e \sqsubseteq \seqPQ$. \qed
\end {proof}

@@

%\EPQOneLarisCoRegular*
\begin{restatable}{lemma}{EPQOneLarisCoRegular}
\label{lemma:EPQ1Lar is co-regular}
$\textit{Auts}_{\textit{1-lar}}$ is $\mathsf{MatchedMaxPriority}^{>}$-complete.
\end{restatable}

@@

\begin{restatable}{lemma}{EPQOneEqualIsCoRegular}
\label{lemma:EPQ1Equal is co-regular}
$\mathsf{MatchedMaxPriority}^{=}$ is co-regular.
\end{restatable}
